---
title: "Yogurt Study — Data Dictionary"
subtitle: "Group A: Understanding the effect of yogurt consumption on the vaginal microbiome after antibiotic treatment"
author: "Workshop Participant"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

## Overview

This dataset comes from a study looking at whether eating yogurt can help
restore the vaginal microbiome after antibiotics. Participants were assigned
to either a **yogurt** group (who ate yogurt daily) or an **unchanged diet**
group (who didn't change their eating habits).

Samples were collected at two time points:

- **Baseline** — before antibiotics
- **After antibiotic** — after a course of antibiotics

There are four CSV files. Below we load them and describe what each one
contains.

```{r}
#| label: setup
#| message: false
library(tidyverse)

# Helper to read a file and show a clean summary
read_and_describe <- function(path) {
  df <- read_csv(path, show_col_types = FALSE)
  cat("\nRows:", nrow(df), "  Columns:", ncol(df), "\n")
  glimpse(df)
  invisible(df)
}

data_dir <- "../data/group_a_yogurt"
```

---

## File 1: `00_sample_ids_yogurt.csv`

**What one row represents:** a single sample collected from one participant
at one time point.

Each participant may appear twice (once at baseline, once after antibiotics)
because they contributed one sample at each time point.

```{r}
#| label: load-sample-ids
sample_ids <- read_and_describe(file.path(data_dir, "00_sample_ids_yogurt.csv"))
```

| Column | Type | What it means |
|--------|------|---------------|
| `pid` | character | Participant ID (e.g. `pid_25`). Links to the metadata file. |
| `time_point` | character | When the sample was taken: `baseline` (before antibiotics) or `after_antibiotic` (after antibiotics). |
| `arm` | character | Which group the participant was in: `yogurt` (ate yogurt daily) or `unchanged_diet` (no change). |
| `sample_id` | character | Unique sample ID. Codes starting with `UNCH` are from the unchanged-diet arm; codes starting with `YOG` are from the yogurt arm. This is the key used to link to qPCR and Luminex results. |

---

## File 2: `01_participant_metadata_yogurt.csv`

**What one row represents:** one participant enrolled in the study.

There are 51 unique participants. Each participant appears exactly once here.

```{r}
#| label: load-metadata
metadata <- read_and_describe(file.path(data_dir, "01_participant_metadata_yogurt.csv"))
```

| Column | Type | What it means |
|--------|------|---------------|
| `pid` | character | Participant ID — matches the same IDs in `sample_ids`. |
| `arm` | character | Group assignment: `yogurt` or `unchanged_diet`. |
| `days_since_last_sex` | double | Number of days since the participant last had sexual intercourse at the time of the baseline visit. |
| `birth_control` | character | Type of birth control used: `no hormonal birth control`, `Depoprovera` (an injectable hormonal contraceptive). |
| `age` | double | Participant's age in years. |
| `education` | character | Highest education level completed: `less than grade 9`, `grade 10-12, not matriculated`, `grade 10-12, matriculated`, or `post-secondary`. |
| `sex` | character | All participants are recorded as `female`. |

---

## File 3: `02_qpcr_results_yogurt.csv`

**What one row represents:** the qPCR (quantitative PCR) measurements for
one sample. qPCR is a lab technique that measures how much of a specific
bacterium's DNA is present.

```{r}
#| label: load-qpcr
qpcr <- read_and_describe(file.path(data_dir, "02_qpcr_results_yogurt.csv"))
```

| Column | Type | What it means |
|--------|------|---------------|
| `sample_id` | character | Sample ID — links to `sample_ids`. |
| `qpcr_bacteria` | double | Total bacterial DNA measured, in **copies per mL** (or equivalent unit). A higher number means more bacteria overall. |
| `qpcr_crispatus` | double | DNA measure specifically for *Lactobacillus crispatus*, a "good" bacterium often linked to vaginal health. In **copies per mL**. A value of 0 means it was not detected. |
| `qpcr_iners` | double | DNA measure specifically for *Lactobacillus iners*, another common vaginal *Lactobacillus* species. In **copies per mL**. |

> **Note:** *L. crispatus* is generally considered beneficial for vaginal
> health, while *L. iners* is sometimes associated with a less stable
> microbiome. You may want to compare how these change between baseline
> and after antibiotics across the two arms.

---

## File 4: `03_luminex_results_yogurt.csv`

**What one row represents:** the concentration of **one cytokine** measured
in **one sample**. This is "long" (tidy) format — each sample has 10 rows,
one per cytokine measured.

Cytokines are small proteins released by immune cells that act as chemical
messengers. They help us understand the immune / inflammatory state at the
sampling site.

```{r}
#| label: load-luminex
luminex <- read_and_describe(file.path(data_dir, "03_luminex_results_yogurt.csv"))
```

```{r}
#| label: list-cytokines
cat("Unique cytokines measured:", 
    paste(unique(luminex$cytokine), collapse = ", "))
```

| Column | Type | What it means |
|--------|------|---------------|
| `sample_id` | character | Sample ID — links to `sample_ids`. |
| `cytokine` | character | Name of the cytokine or chemokine being measured. There are 10 different ones (see list above). |
| `conc` | double | Measured concentration, in **pg/mL** (picograms per millilitre). |
| `limits` | character | Whether the concentration was within the detectable range of the assay: `within limits` means it was measurable, `out of range` means it was below or above the assay's detection limits (so the number might not be fully reliable). |

### What each cytokine generally tells you

| Cytokine | Role (simplified) |
|----------|-------------------|
| **IL-1a** | Pro-inflammatory — signals tissue damage or infection |
| **IL-1b** | Pro-inflammatory — one of the first signals in an immune response |
| **IL-6** | Pro- and anti-inflammatory — involved in fever and immune activation |
| **IL-8** | Chemokine — recruits immune cells to the site of infection |
| **IL-10** | Anti-inflammatory — helps calm down immune responses |
| **TNFa** | Pro-inflammatory — a major "alarm" cytokine |
| **IFN-Y** | Pro-inflammatory — important for fighting viral and bacterial infections |
| **IP-10** | Chemokine — attracts immune cells; often high during inflammation |
| **MIG** | Chemokine — similar to IP-10, attracts immune cells |
| **MIP-3a** | Chemokine — helps guide immune cells to tissues |

---

## How the files connect

Here's a visual of how the files relate to each other — a simple relational
"schema":

```{r}
#| label: schema-drawing
#| echo: false
cat("
  01_participant_metadata     00_sample_ids
  ┌──────────────────┐        ┌──────────────────┐
  │ pid (primary key)│◄───────│ pid (foreign key) │
  │                  │        │ sample_id (PK)   │
  └──────────────────┘        └───────┬──────────┘
                                      │
                         ┌────────────┼────────────┐
                         │            │            │
                         ▼            ▼            ▼
              ┌─────────────────┐ ┌─────────────────────┐
              │ 02_qpcr_results │ │ 03_luminex_results  │
              │ sample_id (FK)  │ │ sample_id (FK)      │
              └─────────────────┘ └─────────────────────┘
")
```

**To combine them in R**, you would join on `pid` (for metadata ↔ samples)
or on `sample_id` (for samples ↔ qPCR / Luminex). Here's an example:

```{r}
#| label: example-join
#| eval: false
# Combine sample info with metadata
full_data <- sample_ids %>%
  left_join(metadata, by = "pid") %>%
  left_join(qpcr, by = "sample_id") %>%
  left_join(luminex, by = "sample_id")

head(full_data)
```

---

## Quick exploration ideas

Here are a few small things you might try to get familiar with the data:

```{r}
#| label: explore-1
# How many participants in each arm?
sample_ids |> 
  distinct(pid, arm) |> 
  count(arm)
```

```{r}
#| label: explore-2
# Age range of participants
summary(metadata$age)
```

```{r}
#| label: explore-3
# Which cytokines are most often out of range?
luminex |> 
  count(cytokine, limits) |> 
  filter(limits == "out of range") |> 
  arrange(desc(n))
```

```{r}
#| label: explore-4
# Compare total bacteria between arms and time points
sample_ids %>%
  left_join(qpcr, by = "sample_id") %>%
  group_by(arm, time_point) %>%
  summarise(
    mean_bacteria = mean(qpcr_bacteria, na.rm = TRUE),
    sd_bacteria = sd(qpcr_bacteria, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| label: plot-bacteria
#| fig-width: 8
#| fig-height: 5
#| message: false
# Boxplot: total bacteria by arm and time point
sample_ids %>%
  left_join(qpcr, by = "sample_id") %>%
  ggplot(aes(x = interaction(arm, time_point, sep = "\n"),
             y = qpcr_bacteria, fill = arm)) +
  geom_boxplot(alpha = 0.6) +
  scale_y_log10(labels = scales::label_comma()) +
  labs(
    x = NULL,
    y = "Total bacteria (copies / mL, log scale)",
    title = "Total bacteria by arm and time point",
    fill = "Arm"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold")) +
  scale_fill_manual(values = c("yogurt" = "#E69F00", "unchanged_diet" = "#56B4E9"))
```